---
title: "Do bigger Star Wars characters weigh more?"
author: "Tamás Nagy"
date: today
format:
  html:
    toc: true
    embed-resources: true
execute:
  warning: false
  message: false
editor_options:
  chunk_output_type: console
---

```{r}
#| label: setup

library(dplyr)
library(ggplot2)

theme_set(theme_light())
```

## The question

Body mass generally scales with height, and there is no obvious reason why the
inhabitants of a fictional galaxy should be an exception. The `starwars` data
that ships with dplyr gives us height and mass for most of the characters, so we
can check.

This is a small enough dataset that we can look at every case, which turns out
to matter.

## The data

```{r}
#| label: glimpse

glimpse(starwars)
```

The data covers `r nrow(starwars)` characters. Height and mass are recorded in
centimetres and kilograms, and `r sum(is.na(starwars$mass))` characters are
missing a mass — a little under a third, so the missingness is worth keeping in
mind rather than ignoring.

## Descriptives

```{r}
#| label: summary-table

starwars |>
  group_by(sex) |>
  summarise(
    n           = n(),
    mean_height = mean(height, na.rm = TRUE),
    mean_mass   = mean(mass,   na.rm = TRUE),
    .groups     = "drop"
  ) |>
  mutate(across(where(is.numeric), \(x) round(x, 1)))
```

Without `na.rm = TRUE` every mean would come back as `NA`, which is R telling us
something true but unhelpful: it cannot average numbers it does not have.

The `hermaphroditic` row is a single character, and its mean mass is a long way
from everyone else's. That is worth remembering when we get to the plot.

## Height against mass

```{r}
#| label: fig-scatter
#| fig-cap: "Height and mass for every character with both recorded."

starwars |>
  ggplot(aes(height, mass)) +
  geom_point(alpha = 0.7) +
  labs(x = "height (cm)", y = "mass (kg)")
```

The plot is almost unreadable, and it is one point's fault. A single character
sits at over 1300 kg while everyone else is below 200, so the y axis stretches to
accommodate it and the remaining `r sum(!is.na(starwars$mass)) - 1` characters
are squashed into a band along the bottom.

Whatever the relationship between height and mass is, this figure cannot show
it.

## The outlier

```{r}
#| label: heaviest

starwars |>
  select(name, height, mass, species) |>
  arrange(desc(mass)) |>
  head(3)
```

It is Jabba, who is a large slug, and whose mass is about eight and a half times
that of the next heaviest character. He is not a data entry error — he really is
that heavy — but he is not the same kind of thing as the others.

```{r}
#| label: correlations

r_all <- cor(starwars$height, starwars$mass, use = "complete.obs")

no_jabba <- starwars |> filter(!grepl("Jabba", name))

r_nojabba <- cor(no_jabba$height, no_jabba$mass, use = "complete.obs")

c(all = r_all, without_jabba = r_nojabba) |> round(3)
```

```{r}
#| label: fig-scatter-2
#| fig-cap: "The same plot with Jabba removed."

no_jabba |>
  ggplot(aes(height, mass)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
  labs(x = "height (cm)", y = "mass (kg)")
```

## Conclusion

With every character included, the correlation between height and mass is
`r round(r_all, 2)` — near enough to nothing. With Jabba excluded it is
`r round(r_nojabba, 2)`, a strong positive relationship of roughly the size you
would expect.

One observation out of `r sum(!is.na(starwars$mass))` moves the answer from "no
relationship" to "a strong one". I would report the second figure, because the
question is about how height and mass scale among broadly humanoid characters
and Jabba is not one. But that is a judgment, not a calculation, and the honest
way to present it is to report both numbers and say which I chose — which is
what this paragraph does.

Note that every number in it was computed rather than typed. If a character were
added to the data tomorrow, re-rendering this file would correct the prose.

## Notes for the instructor

- The `na.rm` stumble in the summary table is deliberate and worth letting
  people hit.
- Task 6 asks for interpretation before the outlier is named. Some will spot
  Jabba immediately; others will only say "one point is weird". Both are fine.
- The 0.13 to 0.75 jump is the moment to connect forward to part 3: here the
  decision is easy because you can see everything. In the blinding exercise you
  will have to make the same kind of call without knowing which way it cuts.
