---
title: "Religiosity and well-being: a blind analysis"
author: "YOUR NAME"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    df-print: kable
    embed-resources: true
editor_options:
  chunk_output_type: console
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(dplyr)
library(ggplot2)
library(readr)
library(lme4)

theme_set(theme_light())
```

::: {.callout-important}
## Before you start

Restart R (Session → Restart R, or Ctrl/Cmd+Shift+F10) and make sure your
environment is empty. You have the original MARP data installed on this machine.
The blinding only works if you do not look at it.

This is how blind analysis works in real life too. It is a protocol, not
encryption.
:::

## The hypothesis

> **H1. People who are more religious report higher well-being.**

That is the entire confirmatory question. Everything else in this document is in
service of answering it defensibly.

## The data

```{r}
#| label: load

d_raw <- read_csv("data/processed/marp_blinded.csv", show_col_types = FALSE)

dim(d_raw)
```

Your partner blinded this file. You do not know their seed, and you do not know
which masked country is which.

**What was blinded:** the religiosity variables were scrambled within country, so
the link between religiosity and well-being is gone. Country labels were masked.

**What was not:** well-being, the covariates, the cluster structure, and every
country-level mean. Distributions, missingness and data quality problems are all
exactly as they were in the real data.

::: {.callout-warning}
## Your blind result will be null. That is not a bug.

Scrambling removes the religiosity–well-being association completely, so the
coefficient you estimate here will sit on top of zero no matter what you do.

So what is the point of fitting it? You are not here to find a result. You are
here to commit — to your exclusions, your covariates, your model form and your
treatment of any odd clusters — at a moment when those choices *cannot* be bent
towards an answer you like. When you unblind at the end, every one of those
decisions is already fixed, and you can prove it.

Other blinding schemes leave an effect in place to look at. This one deliberately
does not.
:::

### The variables you need

| Variable | What it is |
|---|---|
| `wb_overall_mean` | **Outcome.** Mean of 18 well-being items (1–5). Higher = better. |
| `rel_mean` | **Predictor.** Mean of 9 religiosity items, each scaled 0–1. Higher = more religious. |
| `country` | **Cluster.** Masked. 24 groups. |
| `age`, `gender`, `ses`, `education` | Candidate covariates. `ses` and `education` are ordinal integers. |
| `attention_check` | Attention check score. |
| `cnorm_mean` | Perceived cultural norm of religiosity. Not needed for H1. |

The composites are built for you — do not rebuild them. The individual items are
in the file if you want them, but you do not need them today.

## Exploratory data analysis

Run all of this. It is deliberately given to you: the skill being practised today
is *reading* these outputs and deciding what to do, not writing them.

### Overview

```{r}
#| label: eda-overview

d_raw |>
  select(wb_overall_mean, rel_mean, age, ses, education, attention_check) |>
  summary()
```

```{r}
#| label: eda-missing

d_raw |>
  summarise(across(everything(), ~ sum(is.na(.x)))) |>
  tidyr::pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  filter(n_missing > 0) |>
  arrange(desc(n_missing))
```

### Categorical variables

```{r}
#| label: eda-cat

count(d_raw, gender)
count(d_raw, attention_check)
count(d_raw, sample_type)
```

### Distributions

```{r}
#| label: fig-dist
#| fig-height: 3
#| fig-cap: "Outcome and predictor."
#| warning: false

d_raw |>
  select(wb_overall_mean, rel_mean, age) |>
  tidyr::pivot_longer(everything()) |>
  ggplot(aes(value)) +
  geom_histogram(bins = 40) +
  facet_wrap(~ name, scales = "free") +
  labs(x = NULL, y = NULL)
```

### The clusters

This is the most informative plot in the document. Read it carefully.

```{r}
#| label: fig-country
#| fig-height: 5
#| fig-cap: "Mean well-being by (masked) country, with 95% CI. Point size = sample size."

d_raw |>
  group_by(country) |>
  summarise(
    n    = n(),
    mean = mean(wb_overall_mean, na.rm = TRUE),
    se   = sd(wb_overall_mean, na.rm = TRUE) / sqrt(sum(!is.na(wb_overall_mean))),
    .groups = "drop"
  ) |>
  mutate(country = forcats::fct_reorder(country, mean)) |>
  ggplot(aes(x = mean, y = country)) +
  geom_errorbarh(aes(xmin = mean - 1.96 * se, xmax = mean + 1.96 * se), height = 0) +
  geom_point(aes(size = n)) +
  scale_size_area(max_size = 4) +
  labs(x = "mean well-being", y = NULL, size = "n")
```

```{r}
#| label: eda-country-table

d_raw |>
  group_by(country) |>
  summarise(
    n         = n(),
    wb_mean   = mean(wb_overall_mean, na.rm = TRUE),
    wb_sd     = sd(wb_overall_mean,   na.rm = TRUE),
    rel_mean_ = mean(rel_mean, na.rm = TRUE),
    rel_sd    = sd(rel_mean,   na.rm = TRUE),
    .groups   = "drop"
  ) |>
  arrange(wb_mean) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))
```

::: {.callout-note}
## Question 1

Is there anything in the last two outputs you would want to act on before
fitting a model? Write your answer down *now*, before you go further.
:::

## Your cleaning decisions

Fill in the body of this function. Everything you decide about which cases to
keep and how to code variables goes **here and nowhere else**.

Why a function: at the end of the session you will run the identical analysis on
the unblinded data. If your decisions live in one function, that re-run uses
exactly the choices you made while blind, and you can prove it. If they are
scattered through the document, they will drift.

```{r}
#| label: clean

clean_data <- function(d) {
  d |>
    # TODO: which participants should be excluded, if any?

    # TODO: any variables that need recoding?

    # TODO: any implausible values to handle?

    # Standardised versions, so coefficients are comparable. Leave this in.
    mutate(
      rel_z       = as.numeric(scale(rel_mean)),
      age_z       = as.numeric(scale(age)),
      ses_z       = as.numeric(scale(ses)),
      education_z = as.numeric(scale(education))
    )
}

d <- clean_data(d_raw)

c(before = nrow(d_raw), after = nrow(d))
```

## The model

Participants are nested in countries, so a model that ignores country will
understate the uncertainty. Fit something that respects that.

Three levels of hint. Take the lowest one that gets you moving — there is no
prize for struggling.

::: {.callout-tip collapse="true"}
## Hint 1 — what to fit, in words

Predict well-being from religiosity, adjusting for the demographic covariates,
and allow each country to have its own baseline level of well-being.
:::

::: {.callout-tip collapse="true"}
## Hint 2 — the specification

```
outcome    wb_overall_mean
predictor  rel_z
covariates age_z, gender, ses_z, education_z
random     intercept, varying by country
function   lmer(), from the lme4 package
```
:::

::: {.callout-tip collapse="true"}
## Hint 3 — the call

```r
m_blind <- lmer(
  wb_overall_mean ~ rel_z + age_z + gender + ses_z + education_z + (1 | country),
  data = d
)

summary(m_blind)
```
:::

```{r}
#| label: model

# TODO: fit your model, call it m_blind


```

## Diagnostics

Given to you as well. Run them and interpret.

```{r}
#| label: fig-resid
#| fig-height: 4
#| fig-cap: "Residuals against fitted values, coloured by masked country."
#| eval: false

mf <- model.frame(m_blind)

tibble::tibble(
  fitted  = fitted(m_blind),
  resid   = resid(m_blind),
  country = mf$country
) |>
  ggplot(aes(fitted, resid, colour = country)) +
  geom_point(alpha = 0.3, size = 0.7, show.legend = FALSE) +
  geom_hline(yintercept = 0) +
  labs(x = "fitted", y = "residual")
```

```{r}
#| label: fig-ranef
#| fig-height: 5
#| fig-cap: "Country intercepts: how far each country sits from the overall mean."
#| eval: false

re <- ranef(m_blind)$country
tibble::tibble(country = rownames(re), intercept = re[, 1]) |>
  mutate(country = forcats::fct_reorder(country, intercept)) |>
  ggplot(aes(intercept, country)) +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_point() +
  labs(x = "country intercept", y = NULL)
```

::: {.callout-note}
## Question 2

One cluster does not look like the others. Before you decide what to do about
it, consider: you cannot tell whether this is a real difference between
populations or a problem with how the data was collected there.

- What would make you **exclude** it?
- What would make you **keep** it?
- Is there a third option?

You are making this call without knowing which country it is, and without
knowing whether excluding it helps or hurts H1. That is the point of the
exercise. Write your reasoning in `blinding_log.md` **now**, while you still
cannot know the answer.
:::

## Robustness

Refit without the cluster you flagged, and compare. Whatever you decided above,
report both — a decision you can show is robust is worth more than a decision you
merely defend.

```{r}
#| label: robustness

# TODO: refit excluding the flagged cluster, call it m_blind_excl


```

## Unblinding

Stop here until you have written down your decision and your expected result.

**Predict, in writing:** what will the religiosity coefficient be? Positive,
negative, near zero? Will dropping the odd cluster change it?

Now run it.

```{r}
#| label: unblind
#| eval: false

library(vazul)
data(marp)

d_true <-
  marp |>
  mutate(
    rel_mean   = rowMeans(across(rel_1:rel_9), na.rm = TRUE),
    cnorm_mean = rowMeans(across(c(cnorm_1, cnorm_2)), na.rm = TRUE)
  ) |>
  clean_data()            # <- your decisions, unchanged

m_true <- lmer(
  formula(m_blind),       # <- your model, unchanged
  data = d_true
)

# Pull the fixed effects out of a model, no extra packages needed.
fixed_effects <- function(model, label) {
  co <- summary(model)$coefficients
  tibble::tibble(
    term     = rownames(co),
    estimate = co[, 1],
    se       = co[, 2],
    t        = co[, 3],
    data     = label
  )
}

bind_rows(
  fixed_effects(m_blind, "blinded"),
  fixed_effects(m_true,  "unblinded")
) |>
  tidyr::pivot_wider(names_from = data, values_from = c(estimate, se, t)) |>
  mutate(across(where(is.numeric), \(x) round(x, 4)))
```

Which country was it?

```{r}
#| label: reveal
#| eval: false

d_true |>
  group_by(country) |>
  summarise(n = n(), wb = mean(wb_overall_mean, na.rm = TRUE), .groups = "drop") |>
  arrange(wb) |>
  head(5)
```

## Reflection

Answer in a few sentences each, in `blinding_log.md`:

1. Was your prediction right?
2. Did the estimate move when you excluded the flagged cluster? Would you have
   made the same call if you had known which country it was, and known what
   excluding it did to your p-value?
3. Your partner made different blinding choices than you would have. Did anything
   about the file they sent limit what you could check?
4. What would you have to add to this workflow to run a real blind analysis on
   your own data next month?
