---
title: "TODO: give your report a title"
author: "TODO: your name"
date: today
format:
  html:
    toc: true
    embed-resources: true
editor_options:
  chunk_output_type: console
---

<!--
HOW THIS WORKS

Eight tasks. Some ask for code, some ask you to write prose, and a couple ask
for both. Work down the file in order and render whenever you like
(the Render button, or Ctrl/Cmd+Shift+K).

The alternation between code and text is the point. A report is an argument
with evidence in it, not a script with comments.

Delete these instruction comments and the "Task N" headings as you go -- by the
end this should read as a report, not as a worksheet.
-->

```{r}
#| label: setup
#| message: false
#| warning: false

library(dplyr)
library(ggplot2)

theme_set(theme_light())
```

## Task 1 (text)

Fill in `title` and `author` in the YAML above. The title should be the question
you are answering, not "Assignment 2".

Then write two or three sentences here introducing the question: we have data on
Star Wars characters, and we want to know whether taller ones weigh more.

*Your text here.*

## Task 2 (code)

The `starwars` data comes with dplyr, so there is nothing to load from disk.

Have a look at it — `glimpse()` is a good choice. Give the chunk a label.

```{r}
#| label: todo-task-2
# your code here

```

## Task 3 (text and inline code)

Inline code lets you put a computed value inside a sentence. The pattern is a
backtick, then `r`, then the expression, then a closing backtick.

Here is one done for you. Render the document and see what it becomes:

> The data covers `r nrow(starwars)` characters.

Now write your own sentence below, reporting how many characters are missing a
recorded mass. `sum(is.na(starwars$mass))` will give you the number.

*Your sentence here.*

## Task 4 (code)

Make a small summary table: for each `sex`, the number of characters and their
mean height and mean mass.

`group_by()` and `summarise()` will do it. One thing to watch: the means will
come out as `NA` unless you tell `mean()` what to do about missing values.

```{r}
#| label: todo-task-4
# your code here

```

## Task 5 (code)

Plot height on the x axis and mass on the y axis, one point per character.

Give the chunk a label and a figure caption, using the `fig-cap` option.

```{r}
#| label: todo-task-5
#| fig-cap: "TODO: write a caption"
# your code here

```

## Task 6 (text)

Look at your plot and write a short paragraph describing what you see.

There is no right answer here, and that is deliberate — this is the part of a
report that only you can write. Say what the pattern looks like, and mention
anything that strikes you as needing a decision.

*Your paragraph here.*

## Task 7 (code)

Something in that plot is distorting the picture.

Find out which character it is, then compute the correlation between height and
mass twice: once with everyone, and once with that character excluded.

`cor()` needs `use = "complete.obs"` to cope with the missing values.

```{r}
#| label: todo-task-7
# your code here

```

## Task 8 (text and inline code)

Write a closing paragraph reporting both correlations, using inline code for the
two numbers rather than typing them in.

`round()` is your friend here — two decimal places is plenty.

Say which of the two you would report in a paper, and why.

*Your paragraph here.*

---

## When you are done

1. Render one more time and check it reads as a report.
2. Render it into the `docs/` folder so GitHub Pages can find it.
3. Commit and push.
4. Open your Pages URL from part 1 and see your report there.

::: {.callout-tip collapse="true"}
## Stretch, if you have time

- Hide the code in one chunk with `#| echo: false`, so only its output shows.
- Move `message: false` and `warning: false` out of the setup chunk and into the
  YAML so they apply to the whole document. The syntax is:

```yaml
execute:
  warning: false
  message: false
```
:::
