---
title: "Blind analysis — worked solution"
subtitle: "Instructor reference. Do not hand out before the debrief."
format:
  html:
    toc: true
    embed-resources: true
editor_options:
  chunk_output_type: console
---

```{r}
#| label: setup
#| message: false

library(dplyr)
library(ggplot2)
library(readr)
library(lme4)
library(vazul)

theme_set(theme_light())
```

## What participants should find

Three things, in roughly this order of difficulty.

1. **The attention check.** `attention_check` is not binary — it runs 1–7, and
   only `1` means passed. 340 participants (3%) fail. Filtering is the
   conventional choice and matches the original MARP protocol.
2. **Implausible ages.** Eight participants are aged 0–4. The sample is large
   enough that it makes no practical difference, which is itself a useful thing
   to demonstrate: not every anomaly needs action.
3. **The odd cluster.** One masked country sits far below all the others on
   well-being. This is Japan.

## Load a blinded file

```{r}
#| label: load

d_raw <- read_csv("data/processed/marp_blinded.csv", show_col_types = FALSE)
```

## The reveal, ahead of time

```{r}
#| label: truth

data(marp)

marp |>
  group_by(country) |>
  summarise(n = n(),
            wb = mean(wb_overall_mean, na.rm = TRUE),
            .groups = "drop") |>
  arrange(wb) |>
  head(5) |>
  mutate(wb = round(wb, 3))
```

Japan's mean well-being is **3.00**. The next lowest country is **3.49**. The gap
between Japan and the rest is larger than the gap spanning the other 23
countries combined.

In the multilevel model, Japan's country intercept is about **−0.40**, against
**−0.14** for the next most extreme. It is not a borderline case; everyone should
see it.

## Cleaning

```{r}
#| label: clean

clean_data <- function(d) {
  d |>
    filter(attention_check == 1) |>
    mutate(
      gender = recode(gender, man = "Male", woman = "Female", other = "Other"),
      age    = if_else(age < 15 | age > 95, NA_real_, as.numeric(age))
    ) |>
    mutate(
      rel_z       = as.numeric(scale(rel_mean)),
      age_z       = as.numeric(scale(age)),
      ses_z       = as.numeric(scale(ses)),
      education_z = as.numeric(scale(education))
    )
}

d <- clean_data(d_raw)
c(before = nrow(d_raw), after = nrow(d))
```

## Model and diagnostics

```{r}
#| label: model

m_blind <- lmer(
  wb_overall_mean ~ rel_z + age_z + gender + ses_z + education_z + (1 | country),
  data = d
)

round(summary(m_blind)$coefficients, 4)
```

```{r}
#| label: fig-ranef
#| fig-height: 5
#| fig-cap: "Country intercepts under blinding. One cluster is obvious."

re <- ranef(m_blind)$country

tibble::tibble(country = rownames(re), intercept = re[, 1]) |>
  mutate(country = forcats::fct_reorder(country, intercept)) |>
  ggplot(aes(intercept, country)) +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_point() +
  labs(x = "country intercept", y = NULL)
```

```{r}
#| label: odd

odd <- rownames(re)[which.min(re[, 1])]
odd
```

## The decision, and why it is a good teaching case

A mean of 3.00 against 3.49 for the next country can be two very different
things:

- **A real difference.** Japan genuinely reports lower subjective well-being on
  this kind of instrument. This is a well-documented cross-cultural finding.
- **A measurement artefact.** East Asian respondents use the middle of Likert
  scales more and the endpoints less. The low mean would then be partly a
  response-style effect, not a well-being effect.

Nothing in the data settles it. That is the point: participants must make a
defensible call without the answer, and the honest options are *exclude*,
*keep*, or *report both*.

The third is the right answer, and almost nobody offers it unprompted.

## Robustness and unblinding

```{r}
#| label: compare

fixed_effects <- function(model, label) {
  co <- summary(model)$coefficients
  tibble::tibble(term = rownames(co), estimate = co[, 1],
                 se = co[, 2], t = co[, 3], data = label)
}

m_blind_excl <- update(m_blind, data = filter(d, country != odd))

d_true <-
  marp |>
  mutate(
    rel_mean   = rowMeans(across(rel_1:rel_9), na.rm = TRUE),
    cnorm_mean = rowMeans(across(c(cnorm_1, cnorm_2)), na.rm = TRUE)
  ) |>
  clean_data()

m_true      <- lmer(formula(m_blind), data = d_true)
m_true_excl <- update(m_true, data = filter(d_true, country != "Japan"))

bind_rows(
  fixed_effects(m_blind,      "blind"),
  fixed_effects(m_blind_excl, "blind, Japan out"),
  fixed_effects(m_true,       "unblind"),
  fixed_effects(m_true_excl,  "unblind, Japan out")
) |>
  filter(term == "rel_z") |>
  select(data, estimate, se, t) |>
  mutate(across(where(is.numeric), \(x) round(x, 4)))
```

## The punchline for the debrief

The religiosity coefficient is **0.064** with Japan in and **0.064** with Japan
out. The decision that everyone agonised over turns out not to matter at all.

This is the best possible outcome for teaching, and worth saying plainly:

> We can state that the exclusion did not drive the result **because we decided
> before we could see it**. Had we chosen after unblinding, the same number would
> carry no weight — a reader would have no way to tell whether we tried it both
> ways and reported the one we preferred.

Blind analysis does not usually change your estimate. It changes what your
estimate is *worth*.

## Common failure modes

| What they did | What breaks | What `check_blinding.R` says |
|---|---|---|
| Scrambled each column separately | `rel_mean` no longer matches the items | "Composite is still consistent with its items" fails |
| Forgot `.groups = "country"` | Country means flattened, Japan invisible | "Country-level religiosity means are preserved" fails |
| Filtered the attention check before sending | Analyst has nothing to find | "Data has NOT been cleaned for the analyst" fails |
| Masked `country` but kept `gdp` | Mask trivially reversible | caught in `00_preprocess.R`; raise it in the debrief |
| Scrambled `wb_*` too | Nothing left to analyse | "Outcome data is untouched" fails |

If a pair is stuck past the 20-minute mark, hand them
`solutions/03_blinding/01_blind_data_solution.R` and let them get to the
analysis. The analysis is where the lesson is.
